Reasonably Programmable Syntax

### 2017-06-05 popl18/syntax-popl18.tex
progress on major reorg of intro and related work

\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}

The surface syntax of a programming language serves as the user interface between human programmers and the language's semantics. As such, language designers often include various \emph{literal forms} in the surface syntax that are designed to decrease the cognitive cost of constructing and pattern matching over values of selected  types.

For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term  semantically equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the expansion of literal forms is \emph{context independent}. We will return to the concept of context independence throughout this work.} List pattern literals are analagous. Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literal forms for XHTML elements \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} gives examles of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2). 
The desugarings of these literals, not shown, are substantially more verbose and obscure.

Third-party library providers are quite justifiably envious of the privileged status of semantically ordinary types that have been equipped with convenient literal forms by the language designer. After all, it is not difficult to find other examples of types for which programmers have invented specialized syntax. For example, 1) clients of a ``collections'' library might like not just list literals, but also vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library in SML might like Ur/Web style XHTML literals; 3) Ur programmers might further like CSS literals, which Ur/Web does not provide; 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}. Research by \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use string encodings (``stringly-typed programming'') and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}. \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar (e.g. for SQL queries, regexes, file system paths, URLs and others.)

\begin{figure}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-6px}
\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\vspace{-6px}
\label{fig:urweb}
\end{figure}

Given this profusion of possible literal forms, it is not sensible to ask the designers of general-purpose languages to keep up. Instead, there has been considerable interest in mechanisms that allow programmers to introduce new type-specific literal forms in a decentralized manner. Unfortunately, existing mechanisms ...

{This paper addresses this established need by introducing a system of \emph{typed literal macros (TLMs)} that allows programmers to define new literal syntax of nearly arbitrary form within an ML- or Scala-like  language}, i.e. a language with a rich static type system (Sec. 2), support for pattern matching (Sec. 3) and an ML-style module system or Scala-style object system with support for type members and type abstraction (Sec. 5). The examples in this paper are written in VerseML, a new dialect of ML that features TLMs.

@@

%In order to reason about types and binding, client programmers need only have knowledge of 1) the segmentation (e.g. by examining a figure like this presented by a code editor or pretty-printer) and 2) a type annotation on the definition of the applied TLM. No other details about the applied TLM's implementation or the expansion that it generates need to be revealed to the client programmer. In other words, 
TLMs come equipped with useful principles of syntactic abstraction. We will more precisely characterize these abstract reasoning principles as we proceed. For now, to develop some intuitions, consider Figure \ref{fig:K-tsm-example}, which uses TLMs to express the ``unreasonable'' example from Figure \ref{fig:K-dialect}. Without examining the expansion of Line 3, we can reason as follows:

\begin{enumerate}
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for typed expansion of the literal body. 
\item \textbf{(Segmentation)} By examining the segmentation, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way peculiar to this form.
\item \textbf{(Capture)} The system prevents capture, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2 -- it cannot capture an unseen binding introduced in the expansion of Line 3.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, for example, \li{w} is in scope.
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that every expansion it generates will have. We will see an example of a TLM definition in the next section.

@@

\section{Reasoning Criteria and Existing Approaches}
\subsection{Library-Specific Syntax Dialects}\label{sec:syntax-dialects-intro}
One approach available to library providers is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms. Ur/Web's syntax (Fig. 1) is a library-specific dialect of Ur's syntax.

@@

\begin{figure}
\begin{lstlisting}
  val w = compute_w ()
  val x = compute_x ()
  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
\end{lstlisting}
\vspace{-8px}
\caption{An example of unreasonable program text.}
\vspace{-8px}
\label{fig:K-dialect}
\end{figure}

Even putting aside the problem of syntactic conflict, there are questions about just how reasonable sprinkling many possibly unfamiliar library-specific literal forms  throughout a program may be. For example, consider the perspective of a programmer attempting to comprehend (i.e. reason about) the program fragment in Figure \ref{fig:K-dialect}, which is written in a syntax dialect constructed by combining various library-specific dialects of ML's (or Scala's) syntax. If the programmer happens to be familiar with the (intentionally terse) syntax of the stack-based database query processing language K \cite{Whitney:2001:LOR:376284.375783}, then Line 3 might pose few difficulties. If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:

\begin{enumerate}
\item \textbf{(Responsibility)} Which constituent dialect defined the literal form that appears on Line 3?
\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions \li{x} and \li{R} (i.e. expressions of variable form), or parsed in some other way peculiar to this literal form?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2? Or might it capture an unseen binding introduced in the desugaring of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break Line 3, or change its meaning? In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

In short, syntax dialects do not come equipped with  principles of \emph{syntactic abstraction}: if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above) in the usual disciplined manner. This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter unfamiliar library constructs. Forcing the programmer to examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions about segmentation, binding and typing can defeat the ultimate purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}.

@@

\subsection{Another Existing Approach: Term Rewriting}
An alternative approach (and the approach that our mechanism is rooted in) is to leave the context-free syntax of the language unmodified and instead contextually repurpose existing forms using a \emph{term rewriting system}.

For example, OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some term rewriting preprocessor must rewrite \cite{ocaml-manual}. We could attempt to mark a string literal containing Ur/Web-style HTML syntax with a ppx annotation, \li{xml}, as follows:
\begin{lstlisting}[numbers=none]
  [%xml "SSTR<h1>Hello, {[first_name]}!</h1>ESTR"]
\end{lstlisting}
Again, there are problems reasoning modularly and abstractly about such constructions. More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so non-determinism can arise, and it is difficult to reason about \textbf{Responsibility}. It is also difficult to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is arbitrary. For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}

Term-rewriting macro systems address these problems -- the applied macro explicitly identifies the rewriting that wil

Template Haskell / Haskell quasiquotations / early macro systems / reader macros

hygienic macro systems -- problem!

typed macro systems, e.g. Scala. mention MacroML stuff, Herman and Wand 2008/2010 work

string splicing / interpolation

TSLs / TSMs

Non-local rewriting systems suffer from problems analagous to those that plague syntax definition systems. In particular, there can be conflicts because separately defined rewriting rules might attempt to rewrite the same term differently. It can also be difficult to determine which rewriting rule, if any, is responsible for any particular term. Certain conventions have emerged, e.g. OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some presumably corresponding term rewriting preprocessor must rewrite \cite{ocaml-manual}, but a correspondence is not enforced. In any case, it is also difficult to reason abstractly about types and binding given a program subject to a large number of rewriting rules.

Modern \emph{term-rewriting macro systems}, however, have made some progress. In particular:
\begin{enumerate}[nolistsep]
\item Macro systems require that the client explicitly apply the intended rewriting (implemented by a macro) to the term that is to be rewritten, thereby addressing the problems of conflict and determining responsibility. However, it is often unclear whether a given macro  is repurposing the form of a given argument or sub-term thereof, as opposed to treating it parametrically by inserting it unmodified into the generated expansion. This is closely related to the problem of determining a {segmentation} discussed above.
\item \emph{Hygienic} macro systems prevent variable {capture} and enforce {context-independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
\item Although much of the research on macro systems has been for languages in the Lisp tradition \cite{mccarthy1978history}, some progress has also been made on reasoning about types with the design of \emph{typed macro systems} where annotations constrain the macro arguments and the generated expansions. Examples include Scala's macro system \cite{ScalaMacros2013}, formal calculi studied by \citet{DBLP:conf/esop/HermanW08} and \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} (which, it is worth noting, do not give the macro access to the syntax tree of arguments.)
\end{enumerate}

>>>>

\vspace{-4px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
The surface syntax of a programming language serves as the interface between human programmers and the language's semantics. As such, language designers often include various \emph{literal forms} in the surface syntax that are designed to decrease the cognitive cost of constructing and pattern matching over values of certain types defined in the standard library. For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term  semantically equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the expansion of literal forms is \emph{context independent}. We will return to the concept of context independence throughout this work.} List pattern literals are analagous. Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literals for XHTML trees, as defined in Ur/Web's standard library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} shows examples of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2). 
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

Third-party library providers are quite justifiably envious of the special status of these few otherwise ordinary user-defined types. After all, it is not difficult to find other examples of types for which mathematicians, scientists and programmers have invented specialized notations. For example, 1) clients of a ``collections'' library might like not just list literals, but also vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just XHTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}. \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use string encodings and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}. Choosing strings over composite data structures, a.k.a. ``stringly-typed programming'', is quite common in practice -- \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar, e.g. for SQL queries, regexes, file system paths, URLs and others. Literal forms at these types, with support for splicing, would lower the cost of using more compositional encodings. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

\begin{figure}
\vspace{-2px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-10px}
\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\vspace{-7px}
\label{fig:urweb}
\end{figure}

Language designers cannot, of course, hope to accomodate all such types with primitive literal forms. Instead, there has been considerable interest in mechanisms that allow programmers to introduce new literal forms in a decentralized manner. We will cover a number of existing mechanisms, including Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, in Sec. \ref{sec:existing-approaches}. The problem is that these existing mechanisms quickly become \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
For example, consider the perspective of a programmer attempting to comprehend (i.e. reason about) the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a large number of syntax extensions defining user-defined literal forms, some unfamiliar to the programmer, have been imported.

\begin{figure}[h]
\vspace{-9px}
\begin{lstlisting}
  val w = compute_w ()
  val x = compute_x ()
  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
\end{lstlisting}
\vspace{-12px}
\caption{An example of unreasonable program text.}
\vspace{-8px}
\label{fig:K-dialect}
\end{figure}

\noindent
If the programmer happens to be familiar with the intentionally terse syntax of the stack-based database query language K \cite{Whitney:2001:LOR:376284.375783}, then the literal form on Line 3 might be convenient. If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:

\begin{enumerate}
\item \textbf{(Responsibility)} Which syntax extension determines the desugaring of the literal form on Line 3? Might there be many possible desugarings (i.e. syntactic non-determinism) due to conflicts between extensions?
\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 taken as spliced expressions (meaning that they appear directly in the desugaring), or are they parsed in some other way peculiar to this literal form?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding introduced in the desugaring of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break Line 3, or change its meaning? In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. Forcing the programmer to examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions about segmentation, binding and typing can defeat the ultimate purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding and renaming, and types mediate abstraction over function and module implementations.  %From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

This paper introduces a more reasonable approach: typed literal macros (TLMs). TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, whose bodies are left initially unparsed according to the fixed context free syntax of the language (i.e. they are initially parsed like raw string literals.) The applied TLM parses the literal body and generates an expansion during a phase known as typed expansion, which generalizes the usual typing phase. TLMs come equipped with uniquely powerful abstract reasoning principles. We will more precisely characterize these abstract reasoning principles as we proceed. For now, to develop some intuitions, consider Figure \ref{fig:K-tsm-example}, which applies a TLM to express the ``unreasonable'' example from Figure \ref{fig:K-dialect}. Without examining the expansion of Line 3, the client programmer can reason as follows:
\begin{enumerate}
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for typed expansion of the literal body. 
\item \textbf{(Segmentation)} By examining the segmentation, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way peculiar to this form.
\item \textbf{(Capture)} The system prevents capture, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2 -- it cannot capture an unseen binding introduced in the expansion of Line 3.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, for example, \li{w} is in scope.
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that every expansion it generates will have. We will see an example of a TLM definition in the next section.

WE WILL CONSIDER A VARIETY OF TERM REWRITING SYSTEMS, INCLUDING PPX, EARLY MACROS, READER MACROS, TEMPLATE HASKELL, QUASIQUOTES, HYGIENIC MACROS, TYPED HYGIENIC MACROS ALA SCALA, TYPE-SPECIFIC LANGUAGES, IN SEC 6. FOR NOW, LET US JUMP IMMEDIATELY TO OUR SYSTEM OF TYPED LITERAL MACROS AND DEVELOP AN INTUITION FOR HOW THEY ADDRESS THE PROBLEMS RAISED ABOVE. THE XHTML EXAMPLE CAN BE EXPRESSED IN OUR SYSTEM AS FOLLOWS:

BUT THAT DOESNT WORK BECAUSE OF HYGIENE! THERE ARE OTHER PROBLEMS AS WELL,

For example, OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some term rewriting preprocessor must rewrite \cite{ocaml-manual}. We could attempt to mark a string literal containing Ur/Web-style HTML syntax with a ppx annotation, \li{xml}, as follows:
\begin{lstlisting}[numbers=none]
  [%xml "SSTR<h1>Hello, {[first_name]}!</h1>ESTR"]
\end{lstlisting}
Again, there are problems reasoning modularly and abstractly about such constructions. More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so the problems of \textbf{Responsibility} come up. It is also impossible to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is arbitrary. For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}

Term-rewriting macro systems are language-integrated local term rewriting systems that require that the client explicitly apply the intended rewriting, implemented by a macro, to the term that is to be rewritten\todo{readeer macros}. This addresses the issue of \textbf{Responsibility}. However, na\"ive macro systems, like the earliest variants of the Lisp macro system \cite{Hart63a}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and GHC's quasiquotation system \cite{mainland2007s}, do not allow clients to reason abstractly about the remaining issues, again because the expansion that they produce is arbitrary.

\emph{Hygienic} macro systems prevent, or abstractly account for \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}, \textbf{Capture}, and they enforce application-site \textbf{Context Independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}. However, this turns out to make it difficult to use hygienic macros to introduce new literal forms of arbitrary design with support for splicing. Consider again our running XHTML example, which we might try to realize by applying a hygienic macro, \li{xml!}, to a string literal:
\begin{lstlisting}[numbers=none]
  (xml! "SSTR<h1>Hello, {[first_name]}!</h1>ESTR")
\end{lstlisting}
THIS IS BAD BECAUSE FIRSTNAME IS BAD HYGIENE GRAH. MAYBE TALK ABOUT STRING SPLICING AND FRAGMENT SYSTEM STUFF HERE.

PUTTING THAT ASIDE, LET US CONSIDER TYPING. LISP IS DUMB NO TYPES. HERMAN AND WAND USED TYPES ONLY TO ACCOUNT FOR BINDING RELATIONSHIPS. SCALA HAS A TYPED MACRO SYSTEM, BUT THERE IS NO FORMAL ACCOUNT OF IT. MACROML IS ONLY A STAGING MACRO SYSTEM, NO WAY TO PARSE LITERALS. (It is not enough that with Template Haskell / GHC quasiquotation, the generated expansion is typechecked -- to satisfy the \textbf{Typing} criterion, it must be possible to reason abstractly about \emph{what the type of the generated expansion is}.) MAYBE BRING UP TSLS AND TSMS HERE.

TRANSITION TO THE CONTRIBUTIONS OF THIS PAPER: TYPED LITERAL MACROS. REWRITE GENERALIZED LITERAL FORMS, KEEP TRACK OF SPLICED SEGMENTS SO THE HYGIENE MECHANISM IS COMPATIBLE, SEGMENTATION IS ALWAYS POSSIBLE. TYPED, INTEGRATED INTO ML SCALA-LIKE LANGUAGE (SECTION HEADINGS). FORMAL DEFINITION. REASONING PRINCIPLES ARE THEOREMS. THATS HOW WE EVALUATE OUR WORK. (USEFULNESS / USABILITY CONCERNS NEED TO BE ADDRESSED SOMEWHERES)

Although much of the research on macro systems has been for languages in the Lisp tradition \cite{mccarthy1978history}, some progress has also been made on reasoning about types with the design of \emph{typed macro systems} where annotations constrain the macro arguments and the generated expansions. Examples include Scala's macro system \cite{ScalaMacros2013}, formal calculi studied by \citet{DBLP:conf/esop/HermanW08} and \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} (which, it is worth noting, do not give the macro access to the syntax tree of arguments.)

{This paper introducing a system of \emph{typed literal macros (TLMs)} that allows programmers to define new literal syntax of nearly arbitrary form within an ML- or Scala-like  language}, i.e. a language with a rich static type system (Sec. 2), support for pattern matching (Sec. 3) and an ML-style module system or Scala-style object system with support for type members and type abstraction (Sec. 5). The examples in this paper are written in VerseML, a new dialect of ML that features TLMs.

@@

%In order to reason about types and binding, client programmers need only have knowledge of 1) the segmentation (e.g. by examining a figure like this presented by a code editor or pretty-printer) and 2) a type annotation on the definition of the applied TLM. No other details about the applied TLM's implementation or the expansion that it generates need to be revealed to the client programmer. In other words,

@@

\section{Reasoning Criteria and Existing Approaches}\label{sec:existing-approaches}
\subsection{Library-Specific Syntax Dialects}\label{sec:syntax-dialects-intro}
One approach available to library providers is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms. Ur/Web's syntax (Fig. 1) is a library-specific dialect of Ur's syntax.

@@

Even putting aside the problem of syntactic conflict, there are questions about just how reasonable sprinkling many possibly unfamiliar library-specific literal forms  throughout a program may be.

@@

\subsection{Another Existing Approach: Term Rewriting}
